Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-02 demystify.tex
Updates from Overleaf

\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, we would receive an overwhelming number of applications to review.

@@

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. Also, talk about things that adcom members might not know about, e.g., your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

@@

Graduating from top universities \emph{that we recognize} helps. For example, if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

\begin{table}
  \centering
  \footnotesize
  \caption{Different types of PhD funding. All Covered: includes tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\textbf{How to get Fellowship?} You need to apply for them.  Many schools provide such fellowships to attract admitted students.
External fellowships including from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.

Prestigious external fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

>>>>

\paragraph{Why CS depts do not waive application fee?}  Note that some universities do waive application fee, however most do not.  In any case, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, most good CS programs would receive an overwhelming number of applications to review.

@@

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Do talk about things that adcom members might not know about and can help make you different than the other 1000 applicants, e.g.,  your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

This is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

Finally, you should get your SOP reviewed by LoR writers, professors (especially those who have served in adcom), or even graduate PhD students (because they have been through this process).

@@

Graduating from top universities \emph{that we recognize} helps. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

\begin{table}
  \centering
  \footnotesize
  \caption{Different types of PhD funding. All cover tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract admitted students. You likely will not need to do anything and the adcom will recommend such fellowships to admitted students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

You likely need to apply for external fellowships including those from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.
